A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2017-01-31 rigid_bodies.tex
nve review

\subsection{Microcanonical Ensemble: A Review}
algebraic manipulations
Prompted by the introduction of a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble.
Nos\'{e}-Hoover chain thermostat. the non-symplectic character, we used the approximate solution for rotation, as we high quality and non-Hamiltonian dynamics admitos no backward error analyisis. We now will briefly review the key expressions of the metioned paper. First, recall that the translational and rotational degrees of freedom are decoupled. Mathematicaly, the position of a particle $j$,  constituent of a rigid body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the position of the center of mass of the body, $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix and $\vt d_j$ is the position of the particle in the body-fixed frame of reference. Note that unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ are employed to represent the orientations of rigid bodies, which satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\vt q} = 0$.
The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray},
\end{equation*}
which admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation} 
that depend on the quaternion components as
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} aconvenient factorization of matrix $\mt A$. This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rates of change of the quaternion components, which is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
It avoids the use of an ficticious angular velocity component, as Miller~\textit{et al.}\cite{Miller2002}. An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation. Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.

For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), and kinetic energy as follows:

@@

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002}. some algebra manipulations where we defined $\mt \Omega = \mt B(\vt \pi)$, it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$. For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference and thus $\vt F$ is the resultant force on the body. As demonstrated in Abreu, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame. e. Finally, after carrying out the required differentiations of $\mathcal{H}$  and substituting $\tr{\mt B} \vt \pi = - \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
\end{align}
\end{subequations}

we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\eqref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\end{equation}

@@

By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete free rotation, the Liouville operator corresponding to the original case of Eq.~\eqref{eq:EDO_system} (as best noticed in its alternative form, Eq.~\eqref{eq:EDO_system_alternative}) can be written as
\begin{equation}
\label{eq:full_operator_uniaxial}
i\!L = i\!L_t + i\!L_b + i\!L_r,
\end{equation}
where $i\!L_t$ corresponds to the first term in the right-hand of Eq.~\eqref{eq:tras_rot_op}. The factorization we employ here for the Liouville operator in Eq.~\eqref{eq:full_operator} is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

@@

Therefore, if the function $w(\vt x)$ that derives from Eq.~\ref{eq:relation_kappa_w} also satisfies Eq.~\ref{eq:split_kappa_w} for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$. Here, we propose and compare two distinct measure-preserving integrators for a rigid-body system coupled to a single thermostat chain. In the first one, the generalized Liouville operator is written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, as it follows from Eq.~\ref{eq:nhc_system},
\begin{equation}
\label{eq:iL_NHC}
\begin{split}
i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
&+ \sum_{j=1}^{M} \left[\alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}}\right].
\end{split}
\end{equation}

>>>>

\section{Theory}

Prompted by a particular factorization of the rotational matrix, in a previous paper\cite{Abreu_2017} we have reformulated the Hamiltonian equations of motion for rigid bodies, from which we obtained symplectic integrators in the microcanonical (NVE) ensemble. In the following, we skip over the mathematical details of the derivations, which can be found in Abreu~\textit{et al.}\cite{Abreu_2017}, and briefly reviewed some of the final expressions. In the aforementioned paper, we introduced an analytical solution for free-rotation, which would be usefull if any backward error analysis is to be carried out. However, the  Nos\'{e}-Hoover chain thermostat are non-Hamiltonian, so any backward error analysis it is not feasible anymore. Given that, in this paper we employ the approximate solution for the rotational motion , which is also a mathematical simplification of the formulation introduced by Miller~\textit{et al.}\cite{Miller2002}, and exhibits a precision, in terms of the integration erros, almost identical to that achieved using the analytical solution.

\subsection{Microcanonical Ensemble}

When describing the motion of rigid bodies, first recall that the translational and rotational degrees of freedom are decoupled. Mathematicaly, the position of a particle $j$,  constituent of a body, will vary with time as $\vt r_j(t) = \vt r(t) + \tr{[{\mt A}(\vt q(t))]}\vt d_j$, where $\vt r(t)$ is the position of the center of mass of the body, $\tr{\mt A(\vt q(t))}$ is obtained by transposing the rotational matrix and $\vt d_j$ is the position of the particle in the body-fixed frame of reference. Note that unit quaternions $\vt q = \tr {[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ are employed to represent the orientations of rigid bodies, which satisfy the known relation\cite{Goldstein2002} $\tr{\vt q}{\vt q} = \|\vt q\|^2 = q_1^2 + q_2^2 + q_3^2 + q_4^2 = 1$, as well as $\tr{\vt q}{\vt q} = 0$.
The matrix $\mt A$ is computed from a given unit quaternion $\vt q$ by\cite{Allen1989,Miller2002}
\begin{equation*}
\label{eq:A_from_q}
\mt A(\vt q) = \begin{smallarray}{ccc}
q_1^2 + q_2^2 - q_3^2 - q_4^2 & 2(q_2 q_3 + q_1 q_4) & 2(q_2 q_4 - q_1 q_3) \\
2(q_2 q_3 - q_1 q_4) & q_1^2 - q_2^2 + q_3^2 - q_4^2 & 2(q_3 q_4 + q_1 q_2) \\
2(q_2 q_4 + q_1 q_3) & 2(q_3 q_4 - q_1 q_2) & q_1^2 - q_2^2 - q_3^2 + q_4^2  
\end{smallarray},
\end{equation*}
which admits the following factorization:
\begin{equation}
\label{eq:factorization_of_A}
{\mt A} = \tr{\mt B}{\mt C}.
\end{equation} 
that depend on the quaternion components as
\begin{equation}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation}
Note that both $\mt B$ and $\mt C$ depend linearly on the Euler parameters, which makes Eq.~\eqref{eq:factorization_of_A} aconvenient factorization of matrix $\mt A$. This factorization provides a relation between the three-dimensional angular velocity vector $\vt \omega$ and the four-dimensional rates of change of the quaternion components, which is
\begin{equation}
\label{eq:relation_qdot_omega}
\dot{\vt q} = \frac{1}{2} \mt B \vt \omega.
\end{equation}
It avoids the use of an ficticious angular velocity component, as Miller~\textit{et al.}\cite{Miller2002}. An important feature of this relation is that it preserves the unit-norm constraint of a quaternion upon infinitesimal rotation. Note that, for simplicity, we might omit the argument of a matrix-valued function (like $\mt A$, $\vt B$, or $\mt C$) when such argument is $\vt q$.

For a three-dimensional system consisting of $N$ non-collinear rigid bodies that interact via a potential $U(\vt r, \vt q)$, we define the Hamiltonian in terms of its generalized coordinates ($\vt r_i$ and $\vt q_i$) and conjugate momenta ($\vt p_i$ and $\vt \pi_i$), as follows~\cite{Abreu_2017}

@@

The equations of motion are obtained from the Hamiltonian by $\dot{\vt r} = \partial \mathcal{H} / \partial \vt p$, $\dot{\vt p} = -\partial \mathcal{H} / \partial \vt r$, $\dot{\vt q} = \partial \mathcal{H} / \partial \vt \pi$, and $\dot{\vt \pi} = -\partial \mathcal{H} / \partial \vt q$.\cite{Goldstein2002}. In order to Eq.~\eqref{eq:H_NVE} to be easily derivable by $\vt \pi$ we defined $\mt \Omega = \mt B(\vt \pi)$ and it follows that $\tr{\mt B}{\vt \pi} = - \tr{\mt \Omega}{\vt q}$. For a rigid body composed of $n_p$ individual particles, the derivative $\partial U/\partial \vt r$ is obtained by
\[
\diff{U}{\vt r} = -\vt F = -\sum_{j=1}^{n_p} {\vt F_j},
\]
where $\vt F_j$ is the force applied on particle $j$, expressed in the space-fixed frame of reference and thus $\vt F$ is the resultant force on the body. As demonstrated in Abreu, we have
\[
\diff{U}{\vt q} = -2 \mt C \vt \tau = -2 \mt C \sum_{j=1}^{n_p} {\vt \delta_j} \times {\vt F_j},
\]
where $\vt \tau$ is the resultant torque exerted on the body, also expressed in the space-fixed frame. e. Finally, after carrying out the required differentiations of $\mathcal{H}$  and substituting $\tr{\mt B} \vt \pi = - \tr{\mt \Omega} \vt q = 2 {\mt I} \vt \omega$, we obtain the following Hamiltonian system of equations of motion for a rigid body:
\begin{subequations}
\label{eq:EDO_system}
\begin{align}
&\dot{\vt r} = \frac{1}{m} \vt p \\
&\dot{\vt p} = \mt F \\
&\dot{\vt q} = \frac{1}{2} \mt B \vt \omega \label{eq:EDO_q} \\
&\dot{\vt \pi} = \frac{1}{2} \mt \Omega \vt \omega + 2 \mt C \vt \tau \label{eq:EDO_pi}
\end{align}
\end{subequations}

As the columns of $\mt B$ are permutations of the components of $\vt q$, one can write $\mt B = \begin{smallarray}{ccc} \hat{\mt B}_1{\vt q} & \hat{\mt B}_2{\vt q} & \hat{\mt B}_3{\vt q}\end{smallarray}$, where $\hat{\mt B}_1$, $\hat{\mt B}_2$, and $\hat{\mt B}_3$ are the following skew-symmetric permutation matrices:
\[
\hat{\mt B}_1 = \begin{smallarray}{cccc}
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
 0 &  0 &  0 &  1 \\
 0 &  0 & \text{-}1 &  0 \\
\end{smallarray} \quad
\hat{\mt B}_2 = \begin{smallarray}{cccc}
 0 &  0 & \text{-}1 &  0 \\
 0 &  0 &  0 & \text{-}1 \\
 1 &  0 &  0 &  0 \\
 0 &  1 &  0 &  0 \\
\end{smallarray} \quad
\hat{\mt B}_3 = \begin{smallarray}{cccc}
 0 &  0 &  0 & \text{-}1 \\
 0 &  0 &  1 &  0 \\
 0 & \text{-}1 &  0 &  0 \\
 1 &  0 &  0 &  0 \\
\end{smallarray}
\]

we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\eqref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices. In this context, the Hamiltonian of the rigid body can be rewritten as
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\end{equation}

@@

By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ for a complete free rotation, the Liouville operator corresponding to the original case of Eq.~\eqref{eq:EDO_system} (as best noticed in its alternative form, Eq.~\eqref{eq:EDO_system_alternative}) can be written as
\begin{equation}
\label{eq:full_operator_uniaxial}
i\!L = i\!L_t + i\!L_b + i\!L_r,
\end{equation}
The factorization we employ here for the Liouville operator in Eq.~\eqref{eq:full_operator} is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L_{NVE}} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

@@

Therefore, if the function $w(\vt x)$ that derives from Eq.~\ref{eq:relation_kappa_w} also satisfies Eq.~\ref{eq:split_kappa_w} for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$. Here, we propose and compare two distinct measure-preserving integrators for a rigid-body system coupled to a single thermostat chain. In the first one, the generalized Liouville operator is written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:trotter_splitting_NVE} and, as it follows from Eq.~\ref{eq:nhc_system},
\begin{equation}
\label{eq:iL_NHC}
\begin{split}
i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
&+ \sum_{j=1}^{M} \left[\alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}}\right].
\end{split}
\end{equation}
